paper figures

### 2025-02-13 paper_figures.Rmd
take out SM samples

```{r FigS4D, fig.width=2 * fig_scale, fig.height=1.7 * fig_scale}
subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
                   "frac_clusters_SG1","frac_clusters_SG2","frac_clusters_SG3","frac_clusters_SG4")

@@

```{r FigS4E, fig.width=2.25 * fig_scale, fig.height=2 * fig_scale}
tmp <- FB.data.pooled %>%
  dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
  dplyr::filter(Diagnosis %in% c('Control','CVID','CVID-like')) %>%
  dplyr::group_by(status) %>%
  dplyr::mutate(status_R=paste0(status," (R=",signif(cor(alpha_ratio_FACS,alpha_ratio_clusters,use='na.or.complete'),2),")"))
ggplot(tmp, aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
  geom_abline(slope=1, intercept=0, size=.5) +
  #stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
  geom_point() + 
  my_theme() +
  scale_color_manual(values=diagnosis_colors, breaks=unique(tmp$status), labels=unique(tmp$status_R)) +
  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)') +
  theme(legend.position=c(.8,.15),
        legend.box=element_blank(),
        legend.title=element_blank())
```

@@

FB.meta <- FB.data.pooled[FB.samples,] %>%
  dplyr::mutate(Sample_donor=gsub("FB_","FB",Sample_donor),
                dataset=ifelse(grepl('FB',Donor),
                               ifelse(Sample_donor %in% FB_train_donors,'training','testing'),
                               'validation')) %>%
  dplyr::select(Donor,Sample_donor,Diagnosis,dataset,Sex,Age,Timepoint,
                pct_B,pct_IgA,pct_IgG)

@@

hm1 <- Heatmap(t(FB.X),
               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
               cluster_columns=dend,
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               top_annotation=HeatmapAnnotation(df=FB.meta[,c('Diagnosis','Sex','Age',#'pct_B','pct_IgA','pct_IgG',
                                                              'Donor','Timepoint'),drop=FALSE] %>% 
                                                  tibble::remove_rownames() %>%
                                                  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint))),
                                                                Diagnosis=factor(Diagnosis,
                                                                                 levels=c('Control','Subclass','Atypical',
                                                                                          'CVID','ICOS','Kabuki',
                                                                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B'))) %>%
                                                  dplyr::left_join(class_df,by='name') %>%
                                                  dplyr::select(-c(Donor,Timepoint)) %>%
                                                  tibble::column_to_rownames('name'),
                                                show_annotation_name=TRUE,
                                                show_legend=TRUE,
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
                                                                             labels_gp=gpar(fontsize=7),
                                                                             legend_height = unit(5, "mm"),
                                                                             legend_width = unit(3, "mm"),
                                                                             grid_height = unit(3, "mm"),
                                                                             grid_width = unit(3, "mm")),
                                                annotation_name_gp=gpar(fontsize=7),
                                                col=list('Diagnosis'=diagnosis_colors,
                                                         'Class'=class_colors,
                                                         'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
                                                         'Sex'=c('male'='mistyrose2',
                                                                 'female'='lightseagreen'))),
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize = 7),
               row_dend_width = unit(7, "mm"),
               row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=7),
                                         at=c(-2,0,2),
                                         grid_height=unit(4, 'mm'),
                                         grid_width=unit(3, 'mm'),
                                         legend_height=unit(5,'mm'),
                                         legend_width=unit(3,'mm')),
               row_title=NULL,
               name='z-score')

@@

@@

```{r FigS4G, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
         dplyr::filter(dataset!='validation') %>%
         dplyr::left_join(class_df, by='name'), 
       aes(x=PC1,y=PC2, color=Class, shape=dataset)) +
  geom_point(size=1) +
  geom_point(data=pcadat.combined %>% dplyr::filter(dataset=='validation'),
             aes(color=Diagnosis), size=1) +
  geom_line(aes(group=Donor)) +
  my_theme() +
  scale_shape_manual(values=c('testing'=16,'training'=16,'validation'=15), guide='none') +
  scale_color_manual(values=c(class_colors,diagnosis_colors), 
                     breaks=c(names(class_colors),
                              names(diagnosis_colors))) +
  guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,7)),size=2), ncol=4)) +
  theme(legend.key.size=unit(1.5,'mm'),
        legend.position='bottom',
        legend.key.spacing.y=unit(0,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='')
```

@@

```{r Fig4J, fig.width=5.25 * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_clusters_SG1',
                   'mean_cluster_size_downsampled',
                   'n_homology_switch',
                   'n_untemplated_switch',
                   'frac_breaks_intra',
                   'insert_frequency')

>>>>

```{r FigS4D, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale}
subclass.cols <- c("frac_clusters_SA1","frac_clusters_SA2",
                   "frac_clusters_SG1","frac_clusters_SG2","frac_clusters_SG3","frac_clusters_SG4")

@@

```{r FigS4E, fig.width=2.5 * fig_scale, fig.height=1.75 * fig_scale}
tmp <- FB.data.pooled %>%
  dplyr::mutate(alpha_ratio_FACS=pct_IgA/(pct_IgA + pct_IgG)) %>%
  dplyr::filter(Diagnosis %in% c('Control','CVID','CVID-like')) %>%
  dplyr::group_by(status) %>%
  dplyr::mutate(status_R=paste0(status," (R=",signif(cor(alpha_ratio_FACS,alpha_ratio_clusters,use='na.or.complete'),2),")"))
ggplot(tmp, aes(x=alpha_ratio_FACS, y=alpha_ratio_clusters, color=status)) + 
  geom_abline(slope=1, intercept=0, size=.5) +
  #stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.x=.75, label.y=c(.4, .33)) +
  geom_point() + 
  my_theme() +
  scale_color_manual(values=diagnosis_colors, breaks=unique(tmp$status), labels=unique(tmp$status_R)) +
  labs(x='FACS', y='SWIBRID', title='IgA / (IgA + IgG)') +
  theme(legend.position=c(.8,.15),
        legend.box=element_blank(),
        legend.title=element_blank())
```

@@

FB.meta <- FB.data.pooled[FB.samples,] %>%
  dplyr::mutate(Sample_donor=gsub("FB_","FB",Sample_donor),
                dataset=ifelse(grepl('FB',Donor),
                               ifelse(Sample_donor %in% gsub('_','',FB_train_donors),'training','testing'),
                               'validation')) %>%
  dplyr::select(Donor,Sample_donor,Diagnosis,dataset,Sex,Age,Timepoint,
                pct_B,pct_IgA,pct_IgG)

@@

hm_meta <- FB.meta[,c('Diagnosis','Sex','Age',
'Donor','Timepoint'),drop=FALSE] %>% 
  tibble::remove_rownames() %>%
  dplyr::mutate(name=gsub('t','',gsub('FB_','',paste0(Donor,'_',Timepoint))),
                Diagnosis=factor(Diagnosis,
                                 levels=c('Control','Subclass','Atypical',
                                          'CVID','ICOS','Kabuki',
                                          'AID','IgG4def','NFKB2','PI3K','TNFRSF13B'))) %>%
  dplyr::left_join(class_df,by='name') %>%
  dplyr::select(-c(Donor,Timepoint)) %>%
  tibble::column_to_rownames('name')

hm1 <- Heatmap(t(FB.X),
               col=colorRamp2(c(-2,0, 2), c("blue",'gray90', "red")),
               cluster_columns=dend,
               cluster_rows=TRUE,
               show_row_names=TRUE,
               show_column_names=TRUE,
               show_row_dend=TRUE,
               top_annotation=HeatmapAnnotation(df=hm_meta,
                                                show_annotation_name=TRUE,
                                                show_legend=TRUE,
                                                simple_anno_size=unit(2,'mm'),
                                                annotation_legend_param=list(title_gp=gpar(fontsize=7),
                                                                             labels_gp=gpar(fontsize=7),
                                                                             legend_height = unit(5, "mm"),
                                                                             legend_width = unit(3, "mm"),
                                                                             grid_height = unit(3, "mm"),
                                                                             grid_width = unit(3, "mm")),
                                                annotation_name_gp=gpar(fontsize=7),
                                                col=list('Diagnosis'=diagnosis_colors,
                                                         'Class'=class_colors,
                                                         'Age'=colorRamp2(c(20,50,80),c("yellow2","orange2","coral3")),
                                                         'Sex'=c('male'='mistyrose2',
                                                                 'female'='lightseagreen'))),
               column_names_gp=gpar(fontsize=4),
               row_names_gp =gpar(fontsize = 7),
               row_dend_width = unit(7, "mm"),
               row_split = columns %>% tibble::column_to_rownames('col') %>% .[selected.cols, 'category'], 
               heatmap_legend_param=list(labels_gp=gpar(fontsize=7),
                                         title_gp=gpar(fontsize=7),
                                         at=c(-2,0,2),
                                         grid_height=unit(4, 'mm'),
                                         grid_width=unit(3, 'mm'),
                                         legend_height=unit(5,'mm'),
                                         legend_width=unit(3,'mm')),
               row_title=NULL,
               name='z-score')

@@

<!-- plot a decision tree using the top features -->

<!-- ```{r FigS4E, fig.width=2.5 * fig_scale, fig.height=1.5 * fig_scale} -->
<!-- library(randomForest) -->
<!-- library(rpart) -->
<!-- library(rpart.plot) -->

<!-- set.seed(2) -->
<!-- tmp <- cbind(human.X[test.here | train.here,],  -->
<!--              droplevels(human.meta[test.here | train.here, 'status', drop=FALSE]), -->
<!--              hm_meta[test.here | train.here,'Class',drop=FALSE]) -->

<!-- rf <- randomForest(tmp[,selected.cols], tmp$status) -->
<!-- #pred <- predict(rf, type='response', as.matrix(human.X[test.here,])) -->

<!-- # auc1 <- list() -->
<!-- # auc2 <- list() -->
<!-- # for (k in seq(2,10)) { -->
<!-- #   vars <- importance(rf) %>%  -->
<!-- #     as.data.frame() %>%  -->
<!-- #     tibble::rownames_to_column('var') %>% -->
<!-- #     dplyr::slice_max(MeanDecreaseGini, n=k) %>% -->
<!-- #     dplyr::pull(var) -->
<!-- #   # cv_model <- cv.glmnet(as.matrix(human.X[train.here,vars]),  -->
<!-- #   #                       human.meta[train.here,'status']!='Control',  -->
<!-- #   #                       alpha = 0, family='binomial') -->
<!-- #  -->
<!-- #   mod <- glmnet(as.matrix(human.X[train.here,vars]),  -->
<!-- #                 human.meta[train.here,'status']!='Control',  -->
<!-- #                 alpha = 0, family = "binomial", -->
<!-- #                 lambda = cv_model$lambda.min) -->
<!-- #  -->
<!-- #   pred_1 <- predict(mod, type = "response", newx=as.matrix(human.X[test.here,vars])) -->
<!-- #   auc1[[k]] <- as.numeric(roc(human.meta[test.here,'status']!='Control', pred_1)$auc) -->
<!-- #    -->
<!-- #   pred_2 <- predict(mod, type = "response", newx=as.matrix(human.X[validation,vars])) -->
<!-- #   auc2[[k]] <- as.numeric(roc(human.meta[validation,'status']!='Control', pred_2)$auc) -->
<!-- # }   -->

<!-- # vars <- importance(rf) %>%  -->
<!-- #     as.data.frame() %>%  -->
<!-- #     tibble::rownames_to_column('var') %>% -->
<!-- #     dplyr::slice_max(MeanDecreaseGini, n=5) %>% -->
<!-- #     dplyr::pull(var) -->
<!-- #  -->
<!-- # dtree <- rpart(status ~ ., data = tmp[, c(vars, 'status')],   method = "class") -->
<!-- #rpart.plot(dtree, type=5, extra=1, box.palette=diagnosis_colors[c('Control','CVID')], tweak=.5) -->

<!-- # rf <- randomForest(tmp[,selected.cols], as.factor(tmp$Class)) -->
<!-- # vars <- importance(rf) %>%  -->
<!-- #     as.data.frame() %>%  -->
<!-- #     tibble::rownames_to_column('var') %>% -->
<!-- #     dplyr::slice_max(MeanDecreaseGini, n=5) %>% -->
<!-- #     dplyr::pull(var) -->
<!-- # dtree <- rpart(Class ~ ., data = tmp[, c(vars, 'Class')],   method = "class") -->
<!-- #rpart.plot(dtree, type=5, extra=1, box.palette=list('Cluster1'='gray20','darkgoldenrod1','forestgreen'), tweak=.5) -->

<!-- rpart_splits <- function(fit, digits = getOption("digits")) { -->
<!--   splits <- fit$splits -->
<!--   if (!is.null(splits)) { -->
<!--     ff <- fit$frame -->
<!--     is.leaf <- ff$var == "<leaf>" -->
<!--     n <- nrow(splits) -->
<!--     nn <- ff$ncompete + ff$nsurrogate + !is.leaf -->
<!--     ix <- cumsum(c(1L, nn)) -->
<!--     ix_prim <- unlist(mapply(ix, ix + c(ff$ncompete, 0), FUN = seq, SIMPLIFY = F)) -->
<!--     type <- rep.int("surrogate", n) -->
<!--     type[ix_prim[ix_prim <= n]] <- "primary" -->
<!--     type[ix[ix <= n]] <- "main" -->
<!--     left <- character(nrow(splits)) -->
<!--     side <- splits[, 2L] -->
<!--     for (i in seq_along(left)) { -->
<!--       left[i] <- if (side[i] == -1L) -->
<!--                    paste("<", format(signif(splits[i, 4L], digits))) -->
<!--                  else if (side[i] == 1L) -->
<!--                    paste(">=", format(signif(splits[i, 4L], digits))) -->
<!--                  else { -->
<!--                    catside <- fit$csplit[splits[i, 4L], 1:side[i]] -->
<!--                    paste(c("L", "-", "R")[catside], collapse = "", sep = "") -->
<!--                  } -->
<!--     } -->
<!--     cbind(data.frame(var = rownames(splits), -->
<!--                      type = type, -->
<!--                      node = rep(as.integer(row.names(ff)), times = nn), -->
<!--                      ix = rep(seq_len(nrow(ff)), nn), -->
<!--                      left = left), -->
<!--           as.data.frame(splits, row.names = F)) -->
<!--   } -->
<!-- } -->

<!-- train.here <- (FB.meta$dataset=='training') & !(FB.meta[,'Diagnosis'] %in% c('Subclass','Atypical'))  -->
<!-- test.here <- (FB.meta$dataset=='testing') & !(FB.meta$Donor %in% FB_train_donors) -->

<!-- FB.dat <- FB.data.pooled %>% -->
<!--   dplyr::mutate(status=as.factor(ifelse(Diagnosis=='Control','Control','CVID'))) -->

<!-- tmp <- lapply(selected.cols, function(col) { -->
<!--   mod <- rpart(status ~ ., data=FB.dat[train.here,c(col,'status')], -->
<!--                method='class') -->
<!--   pred <- predict(mod, newdata=FB.dat[test.here, col, drop=FALSE], type='class') -->
<!--   acc <- as.numeric(confusionMatrix(pred, FB.dat[test.here,'status'])$overall['Accuracy']) -->
<!--   split <- rpart_splits(mod)[1, 'left'] -->
<!--   c(col=col, acc=acc, split=split) -->
<!-- }) -->

<!-- ggplot(do.call(rbind, tmp) %>% -->
<!--          as.data.frame() %>% -->
<!--          dplyr::mutate(acc=as.numeric(acc)) %>% -->
<!--          dplyr::left_join(columns, by='col') %>% -->
<!--          #dplyr::group_by(category) %>% -->
<!--          #dplyr::slice_max(acc, n=3, with_ties=FALSE) %>% -->
<!--          #dplyr::ungroup() %>% -->
<!--          dplyr::slice_max(acc, n=10, with_ties=FALSE) %>% -->
<!--          dplyr::arrange(acc) %>% -->
<!--          dplyr::mutate(col=factor(gsub("_downsampled","",col),  -->
<!--                                   levels=gsub('_downsampled','',col))) %>% -->
<!--          rowwise() %>% -->
<!--          dplyr::mutate(spl=paste(str_split(split, ' ')[[1]][1], signif(as.numeric(str_split(split, ' ')[[1]][2]),3))), -->
<!--        aes(x=col, y=acc, fill=category)) +  -->
<!--   geom_col() +  -->
<!--   geom_text(aes(x=col, y=.05, label=spl), size=2, hjust=0) + -->
<!--   #geom_hline(yintercept=mean(unlist(sample.CV)), linewidth=.5, linetype='dashed') + -->
<!--   my_theme() + -->
<!--   coord_flip() + -->
<!--   scale_fill_manual(values=cat_cols) + -->
<!--   labs(y='accuracy', x='', title='prediction using single feature') + -->
<!--   theme(legend.position='none') -->

<!-- ``` -->

@@

```{r FigS4G, fig.width=2.5 * fig_scale, fig.height=2 * fig_scale}
ggplot(pca_hm %>% tibble::rownames_to_column('name') %>%
         dplyr::filter(dataset!='validation') %>%
         dplyr::left_join(class_df, by='name'), 
       aes(x=PC1,y=PC2, color=Class, shape=dataset)) +
  geom_point(size=1) +
  geom_point(data=pcadat.combined %>% dplyr::filter(dataset=='validation'),
             aes(color=Diagnosis), size=1) +
  geom_line(aes(group=Donor)) +
  my_theme() +
  scale_shape_manual(values=c('testing'=16,'training'=16,'validation'=15), guide='none') +
  scale_color_manual(values=c(class_colors,diagnosis_colors), 
                     breaks=c(names(class_colors),
                              names(diagnosis_colors))) +
  guides(color=guide_legend(override.aes = list(shape=c(rep(16,3),rep(15,6)),size=2), ncol=4)) +
  theme(legend.key.size=unit(1.5,'mm'),
        legend.position='bottom',
        legend.key.spacing.y=unit(0,'mm'),
        aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2) +
  labs(x='PC 1', #paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'),
       y='PC 2', #paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'),
       color='')
```

@@

```{r Fig4J, fig.width=5.25 * fig_scale, fig.height=1.7 * fig_scale}
features.here <- c('frac_breaks_SA_SA',
                   'mean_cluster_size_downsampled',
                   'n_homology_switch',
                   'n_untemplated_switch',
                   'frac_breaks_intra',
                   'insert_frequency')
